---
title: "code"
author: "DS2"
date: "5/6/2022"
output: github_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,dpi=300)
library(tidyverse)
library(caret)
library(patchwork)
library(mgcv)
library(earth)
library(corrplot)
library(vip)
library(ggpubr)

```


## Part 0 - Data Preprocessing

```{r joining datasets}
df_salary = read_csv("NBA_season2122_player_salary.csv") %>%
  janitor::clean_names() %>%
  select(Player=x2,Team=x3,Salary=salary_4) %>%
  na.omit()

df_salary = df_salary[-1,]

df_stats = read_csv("NBA_season2122_player_stats.csv") %>%
  rename(Team=Tm) %>%
  select(-Rk)

df_players = inner_join(x=df_salary,y=df_stats,by=c("Player","Team")) %>% 
  janitor::clean_names() %>% 
  distinct()

df_players = df_players %>% 
  arrange(player,desc(g)) %>% 
  distinct(player,.keep_all = TRUE)

# Removed variables with missing data and resulted from division of other variables
df_players = df_players %>% 
  select(-x3p_percent, -ft_percent, -fg_percent,-x2p_percent,-e_fg_percent)

# The final generated dataset for use: df_player.
```


```{r data cleaning}
# Convert salary from characters to numbers.
# Convert categorical variables to factors

df_players = df_players %>% 
  separate(salary,into = c("symbol", "salary"),1) %>% 
  select(-symbol)%>% 
  mutate(salary = as.numeric(salary)/1000000,
         team = factor(team),
         pos = factor(pos)) %>% 
  relocate(salary, .after = last_col())

colnames(df_players) = c("player", "team", "position", "age", "game","game_starting" ,"minute","field_goal", "fg_attempt", "x3p", "x3p_attempt" ,"x2p", "x2p_attempt",   "free_throw",   "ft_attempt", "offensive_rb", "defenssive_rb",  "total_rb" ,   "assistance" ,   "steal" , "block",    "turnover",  "personal_foul", "point", "salary")

```

## Part 1 - Exploratory Analysis 

### Univariate Analysis

Distributions of the two categorical variables, `team` and `position`.

```{r}
par(mfrow=c(1,2))
plot_team = ggplot(df_players) + geom_bar(aes(team)) + 
  scale_x_discrete(guide = guide_axis(check.overlap = TRUE))+ theme_bw()
plot_team
plot_position = ggplot(df_players) + geom_bar(aes(position)) + theme_bw()
plot_position
```

Distributions of other numeric variables. 

```{r dpi=300}
plot_data_column = function (data, column) {
    ggplot(data, aes_string(x = column)) +
        geom_histogram(bins=15) +
        xlab(column) + theme_bw(base_size = 10)
}

histograms <- lapply(colnames(df_players)[4:24], 
                       plot_data_column, data = df_players)

figure_a = ggarrange(plotlist = histograms[1:9], 
          ncol = 3, nrow = 3)

annotate_figure(figure_a, 
                top = text_grob("Histograms of Predictive Variables (Group A)", 
                                face = "bold", size = 15))


figure_b = ggarrange(plotlist = histograms[10:18], 
          ncol = 3, nrow = 3)
annotate_figure(figure_b, 
                top = text_grob("Histograms of Predictive Variables (Group B)", 
                                face = "bold", size = 15))

figure_c = ggarrange(plotlist = histograms[19:21],
          ncol = 3, nrow = 3)
annotate_figure(figure_c, 
                top = text_grob("Histograms of Predictive Variables (Group C)", 
                                face = "bold", size = 15))
```


### Correlation Analysis

```{r}

df_corr_1 = df_players %>% 
  select(-player, -team,-position)

corrplot(cor(df_corr_1),type = "lower")

```


### Analyzing trends in data

From numeric variables, we found that `stl`,`x3p`, `age`,`gs` seem to have some non-linear trends.

```{r, fig.height=4}
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(0, 0, 0, 1) 
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(1, .1, .1, 1) 
theme1$plot.line$lwd <- 2 
theme1$strip.background$col <- rgb(.0, .2, .6, .2) 
trellis.par.set(theme1)

df_features = df_players[4:24]
featurePlot(x = df_features, 
            y = df_players$salary,
            plot = "scatter",
            # span = .5,
            labels = c("Predictors","Player Salary (Millions)"), 
            type = c("p", "smooth"), 
            layout = c(7, 3))
```

From categorical variable `position`, extremely high values in salary show in all positions and some teams.

```{r fig.height=4}
df_players %>% 
  mutate(
    pos = fct_reorder(position,salary)
  ) %>% 
  ggplot(aes(x = position, y = salary, group = pos, fill = pos))+
  geom_boxplot() + theme_bw()

```


